Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-01-22 PAPER.tex
covariance language edits throughout section 2

One choice for the weighting matrix $\textbf{R}$ used in power spectrum analysis is an inverse covariance matrix. This type of weighting is attractive for power spectrum analyses because it yields the smallest possible error bars on a measurement. Said differently, it gives the minimum variance estimate of the power spectrum of a signal (\citealt{tegmark_et_al1997a}; \citealt{bond_et_al1998}). In \citet{liu_tegmark2011}, it was shown that inverse covariance weighting also serves as a way to down-weight foregrounds since foregrounds have large covariances in frequency, motivating our use of the weighting for previous PAPER analyses.

@@

To begin to understand the lossy behavior of this result, we can closely study our covariance eigenspectrum, shown in Figure \ref{fig:toy_sigloss2}. If the covariance is estimated from our data, its eigenspectrum differs from the eigenspectrum of the true covariance $\textbf{C}$, and this difference has consequences on our result. An eigenspectrum ranks the eigenvalues of a matrix from 
highest to lowest and can be thought of as a spectrum of weights that are given to each spectral mode in the data. In other 
words, the eigenvalues encode the strength of different shapes in the dataset. The eigenspectrum of the identity matrix $
\textbf{I}$ is flat (all $1$'s) because it gives equal weighting to all modes. This is usually not the case for a covariance matrix, 
for which a sloped eigenspectrum means that modes are given different weights in OQE power spectrum estimation. The modes with the highest eigenvalues are 
down-weighted the most.

@@

Because $\widehat{\textbf{C}}$ is estimated from the data, its eigenvectors and eigenvalues are strongly coupled to the particular data realization that was used to compute it. The strongest mode of $\widehat{\textbf{C}}$ (highest eigenvalue) describes the sinusoid foreground mode in the toy model (the first peak in Figure \ref{fig:toy_sigloss2}). In Figure \ref{fig:toy_sigloss12} we show the covariances of our toy model datasets along with inverse covariance weighted data. The foreground sinusoid is clearly visible in $\widehat{\textbf{C}}_{FG}$.

In general, the strongest eigenmodes of $\widehat{\textbf{C}}$ typically describe bright foregrounds --- the most prominent shapes in a dataset. For these `strong' modes, where foregrounds outshine the EoR signal, down-weighting is beneficial. In some sense, we desire signal loss, if by `signal' we mean `foregrounds.' In this case it is okay for the `strong' eigenmodes to be coupled to the data in a way that produces loss. For our toy model, the successful suppression of the foreground mode is demonstrated in Figure \ref{fig:toy_sigloss2} by the missing foreground peak in the weighted power spectrum estimate (green).

@@

The danger of an empirically estimated covariance matrix comes mostly from not being able to describe the `weak' eigenmodes of $
\textbf{C}$ accurately, for which the EoR signal is brighter than foregrounds. \cc{start editing here later!} In such a case, the empirical covariance believes these modes should be down-weighted more than the true covariance would. For these modes, the error that results from not computing $\textbf{C}$ accurately (due to a finite data sample), has a larger impact on the `weak' modes than the `strong' modes. In other words, mis-estimating $\textbf{C}$ affects EoR-dominated eigenmodes more than FG-dominated modes, and it is primarily the coupling between these mis-estimated modes to the data that leads to signal loss.

Using what we've learned about the eigenspectrum, we can tweak it in a simple way to suppress foregrounds and yield minimal 
signal loss. Recall that our toy model foreground is a sinusoid, so it can be perfectly described by a single eigenmode. Using the full dataset's (foreground plus EoR signal) empirical covariance, we can 
project out the first eigenmode (i.e. use the true eigenvalue of the first eigenmode, but flatten the rest of the spectrum to have eigenvalues of 1), thereby down-weighting the foreground-dominated mode more than the rest of the modes. Hence, we are not using the relative weighting information from all the weaker eigenmodes which have comparatively larger errors. Altering $\widehat{\textbf{C}}$ as such is one specific example of a regularization method for this toy model, in which we are 
changing $\widehat{\textbf{C}}$ in a way that flattens its eigenspectrum, making it more similar to that of $\textbf{I}$. The resulting 
power spectrum estimate for this case is shown in the right plot of Figure \ref{fig:toy_sigloss3}. In this case we recover EoR, 
demonstrating that if we can disentangle the foreground-dominating modes from EoR-dominating modes, we can down-weight 
them with negligible signal loss. There are several other ways to regularize $\widehat{\textbf{C}}$, and we will discuss some in Section 
\ref{sec:otherweight}.

@@

With 
fringe-rate filtering resulting in fewer independent modes, it becomes more difficult for the empirical covariance to estimate the 
true fringe-rate filtered case covariance matrix. This can be quantified by evaluating a convergence metric $\varepsilon(\Chat)$ for the empirical covariance, which we define as
\begin{equation}
\label{eq:converge}
\varepsilon (\Chat) \equiv \sqrt{\frac{\sum_{ij} (\widehat{C}_{ij} - {C}_{ij})^2}{\sum_{ij} {C}_{ij}^2}},
\end{equation}
where $\C$ is true covariance matrix. In Figure \ref{fig:toy_sigloss16} we show this convergence statistic as a function of the number of independent ensemble realizations in one's simulations (horizontal axis) and the number of independent samples in the data following fringe-rate filtering (different curve colors). With more independent time samples in the data, one converges to the true fringe-rate filtered covariance more quickly.

@@

Just as important as the eigenvalues are the eigenvectors of our empirical covariances. In general, different eigenvectors converge to their true forms at different rates. This is illustrated by Figure \ref{fig:toy_sigloss17}, which shows the convergence of eigenvectors in an empirical estimate of a covariance matrix whose true form is a diagonal matrix with eigenvalues spanning four orders of magnitude. We use as a convergence metric $\varepsilon(\widehat{\textbf{v}})$ for the empirical eigenvectors $\widehat{\textbf{v}}$, defined as:
\begin{equation}
\label{eq:converge_eig}
\varepsilon (\widehat{\textbf{v}}) \equiv \sqrt{\sum_{i}^{N_{f}}|\textbf{v}-\widehat{\textbf{v}}|_{i}^2},
\end{equation}
where $N_{f}$ is the number of frequencies ($20$) in the mock data. The eigenmode convergence curves  in Figure \ref{fig:toy_sigloss17} are ranked ordered by eigenvalue, such that ``Eigenmode \#0" illustrates the convergence of the eigenvector with the largest eigenvalue, ``Eigenmode \#1" for the second largest eigenvalue, and so on. One sees that the stronger eigenmodes converge to their true eigenvectors more quickly. With only a small number of realizations, these empirically estimated modes already retain little correlation with the specific realizations of data that were used to form the empirical covariance. As we will see in the next section, using only the strongest eigenmodes, which are less coupled to data realizations, minimizes signal loss. In contrast, the weaker eigenmodes retain more memory of the data realizations, which leads to correlations that induce signal loss. %\acl{Think about whether we need to do the Fourier space version as well.}

@@

The second panel (top right) in Figure \ref{fig:toy_sigloss8} uses a regularization method of setting $\widehat{\textbf{C}} = 
\widehat{\textbf{C}} + \gamma\textbf{I}$, where $\gamma = 5$ (an arbitrary strength 
of $\textbf{I}$ for the purpose of this toy model). By adding the identity matrix, element-wise, we are weighting the diagonal 
elements of the covariance matrix more heavily than those off-diagonal, thereby flattening out its eigenspectrum. If all modes are given 
similar weights, we avoid down-weighting EoR modes more than others. Although there is negligible signal loss, the small green peak at $k\sim3$ represents residual foregrounds that still exist since the shapes encoded in the off-diagonal frequency correlations of the covariance matrix were deemed not as prominent as the diagonal elements using this weighting scheme.

The third panel (bottom left) in Figure \ref{fig:toy_sigloss8} flattens out the eigenspectrum of $\widehat{\textbf{C}}$ a different way - 
by only using the first three eigenmodes. Recalling that our toy model foregrounds can be described entirely by the first eigenmode, this 
method intentionally projects out modes that are EoR-dominated by replacing all but the three highest weights in the 
eigenspectrum with $1$'s (equal weights). Again, flattening the eigenspectrum results in negligible signal loss. However, we do 
not perfectly recover the shape of EoR because we lost information when projecting out certain modes. While this case is illuminating for the toy model, in practice it is not obvious which eigenmodes are FG vs. EoR dominated, so selecting a subset of modes to down-weight is not trivial.

The last regularization scheme we are highlighting here is setting $\widehat{\textbf{C}} = \widehat{\textbf{C}} \circ \textbf{I}$ (element-wise multiplication), or inverse variance weighting (keeping only the diagonal elements of $\widehat{\textbf{C}}$). In the bottom right 
panel of Figure \ref{fig:toy_sigloss8}, we see that this method does not down-weight the foregrounds. For this toy model, 
our foregrounds are spread out in frequency and therefore have non-negligible frequency-frequency correlations. Multiplying by 
the identity, element-wise, results in a diagonal matrix, meaning we are only left with correlation information between the same 
two frequencies. Because we disregard information from all other frequency combinations in this case, we do a poor job 
suppressing the foreground. But because we flattened the eigenspectrum, we also avoid signal loss. Although this method did not successfully recover EoR for this particular simulation, it is important that we show that there 
are many options for estimating a covariance matrix, and some may be more effective than others based on the spectral nature 
of the components in a dataset. 
%One may imagine a situation where a particular systematic is contained to an isolated 
%frequency (such as radio frequency interference or crosstalk). In such a case, preserving only the diagonal elements of $
%\widehat{\textbf{C}}$ would be an effective way of removing this contamination.

>>>>

One choice for the weighting matrix $\textbf{R}$ used in power spectrum analysis is an inverse covariance matrix. This type of weighting is attractive for power spectrum analyses because it yields the smallest possible error bars on a measurement. Said differently, it gives the minimum variance estimate of the power spectrum of a signal (\citealt{tegmark_et_al1997a}; \citealt{bond_et_al1998}). In \citet{liu_tegmark2011}, it was shown that inverse covariance weighting also serves as a way to down-weight some foregrounds since foregrounds have large covariances in frequency, motivating our use of the weighting for previous PAPER analyses.

@@

To begin to understand the lossy behavior of this result, we can closely study our estimated covariance eigenspectrum, shown in Figure \ref{fig:toy_sigloss2}. Since it is estimated from our data, its eigenspectrum differs from the eigenspectrum of the true covariance $\textbf{C}$, and this difference has consequences on our result. An eigenspectrum ranks the eigenvalues of a matrix from 
highest to lowest and can be thought of as a spectrum of weights that are given to each spectral mode in the data. In other 
words, the eigenvalues encode the strength of different shapes in the dataset. The eigenspectrum of the identity matrix $
\textbf{I}$ is flat (all $1$'s) because it gives equal weighting to all modes. This is usually not the case for a covariance matrix, 
for which a sloped eigenspectrum means that modes are given different weights in OQE power spectrum estimation. The modes with the highest eigenvalues are 
down-weighted the most.

@@

Because $\widehat{\textbf{C}}$ is estimated from the data, its eigenvectors and eigenvalues are strongly coupled to the particular data realization that was used to compute it. For example, the strongest mode of $\widehat{\textbf{C}}$ (highest eigenvalue) describes the sinusoid foreground mode in the toy model (the first peak in Figure \ref{fig:toy_sigloss2}). In Figure \ref{fig:toy_sigloss12} we show the covariances of our toy model datasets along with inverse covariance weighted data. The foreground sinusoid is clearly visible in $\widehat{\textbf{C}}_{FG}$.

In general, the strongest eigenmodes of $\widehat{\textbf{C}}$ typically describe bright foregrounds --- the most prominent shapes in a dataset. For these `strong' modes, where foregrounds outshine the EoR signal, down-weighting is beneficial. In some sense, we desire signal loss in this regime, if by `signal' we mean `foregrounds.' In this case it is beneficial for the `strong' eigenmodes to be coupled to the data in a way that produces loss. For our toy model, the successful suppression of the foreground mode is demonstrated in Figure \ref{fig:toy_sigloss2} by the missing foreground peak in the weighted power spectrum estimate (green).

@@

The danger of an empirically estimated covariance matrix comes mostly from not being able to describe the `weak' eigenmodes of $
\textbf{C}$ accurately, for which the EoR signal is brighter than foregrounds. In such a case, the coupling between these modes to the data realization leads to the overfitting and subtraction of the EoR signal. In other words, mis-estimating $\textbf{C}$ for EoR-dominated eigenmodes is more harmful than for FG-dominated modes, and since the `weak' modes of an eigenspectrum are typically EoR-dominated, using this part of the spectrum for weighting is most dangerous.

Using what we've learned about the eigenspectrum, we can tweak it in a simple way to suppress foregrounds and yield minimal 
signal loss. Recall that our toy model foreground is a sinusoid, so it can be perfectly described by a single eigenmode. Using the full dataset's (foreground plus EoR signal) empirical covariance, we can 
project out the first eigenmode (i.e. use the true eigenvalue of the first eigenmode, but flatten the rest of the spectrum to have eigenvalues of 1), thereby down-weighting the foreground-dominated mode more than the rest of the modes. Hence, we are changing the weaker part of the spectrum to be less coupled to the data, limiting the amount of over-fitting that can happen for those modes (i.e. only allowing over-fitting to occur for the first mode).

Altering $\widehat{\textbf{C}}$ as such is one specific example of a regularization method for this toy model, in which we are 
changing $\widehat{\textbf{C}}$ in a way that changes its coupling to the data realization. The resulting 
power spectrum estimate for this case is shown in the right plot of Figure \ref{fig:toy_sigloss3}. In this case we recover EoR, 
demonstrating that if we can disentangle the foreground-dominating modes from EoR-dominating modes, we can down-weight 
them with negligible signal loss. There are several other ways to regularize $\widehat{\textbf{C}}$, and we will discuss some in Section 
\ref{sec:otherweight}.

@@

With 
fringe-rate filtering resulting in fewer independent modes, it becomes more difficult for the empirical covariance to estimate the 
true fringe-rate filtered case covariance matrix. This can be quantified by evaluating a convergence metric $\varepsilon(\Chat)$ for the empirical covariance, which we define as
\begin{equation}
\label{eq:converge}
\varepsilon (\Chat) \equiv \sqrt{\frac{\sum_{ij} (\widehat{C}_{ij} - {C}_{ij})^2}{\sum_{ij} {C}_{ij}^2}},
\end{equation}
where $\C$ is true covariance matrix. In Figure \ref{fig:toy_sigloss16} we show this convergence statistic as a function of the number of independent ensemble realizations in one's simulations (horizontal axis) and the number of independent samples in the data following fringe-rate filtering (different curve colors). With more independent time samples (i.e. more realizations) in the data, one converges to the true fringe-rate filtered covariance more quickly.

@@

Just as important as the eigenvalues are the eigenvectors of our empirical covariances. In general, different eigenvectors converge to their true forms at different rates. This is illustrated by Figure \ref{fig:toy_sigloss17}, which shows the convergence of eigenvectors in an empirical estimate of a covariance matrix whose true form is a diagonal matrix with eigenvalues spanning four orders of magnitude. We use as a convergence metric $\varepsilon(\widehat{\textbf{v}})$ for the empirical eigenvectors $\widehat{\textbf{v}}$, defined as:
\begin{equation}
\label{eq:converge_eig}
\varepsilon (\widehat{\textbf{v}}) \equiv \sqrt{\sum_{i}^{N_{f}}|\textbf{v}-\widehat{\textbf{v}}|_{i}^2},
\end{equation}
where $N_{f}$ is the number of frequencies ($20$) in the mock data. The eigenmode convergence curves  in Figure \ref{fig:toy_sigloss17} are ranked ordered by eigenvalue, such that ``Eigenmode \#0" illustrates the convergence of the eigenvector with the largest eigenvalue, ``Eigenmode \#1" for the second largest eigenvalue, and so on. One sees that the stronger eigenmodes converge to their true eigenvectors more quickly. With only a small number of realizations, these empirically estimated modes already retain little correlation with the specific realizations of data that were used to form the empirical covariance. As we will see in the next section, using only the strongest eigenmodes, which are less coupled to data realizations, minimizes signal loss. In contrast, the weaker eigenmodes retain more memory of the data realizations, which leads to correlations that induce signal loss. Said differently, a steep covariance eigenspectrum can be especially dangerous because it is the `weak' modes that are both EoR-dominated and converge the slowest and are therefore susceptible to the most loss.  %\acl{Think about whether we need to do the Fourier space version as well.}

@@

The second panel (top right) in Figure \ref{fig:toy_sigloss8} uses a regularization method of setting $\widehat{\textbf{C}} = 
\widehat{\textbf{C}} + \gamma\textbf{I}$, where $\gamma = 5$ (an arbitrary strength 
of $\textbf{I}$ for the purpose of this toy model). By adding the identity matrix, element-wise, we are weighting the diagonal 
elements of the covariance matrix more heavily than those off-diagonal. Since the identity component does not know anything about the data realization, it alters the covariance to be less coupled to the data. Although there is negligible signal loss using this regularization, the small green peak at $k\sim3$ represents residual foregrounds that still exist since the shapes encoded in the off-diagonal frequency correlations of the covariance matrix were deemed not as prominent as the diagonal elements using this weighting scheme.

The third panel (bottom left) in Figure \ref{fig:toy_sigloss8} minimizes signal loss a different way - 
by only using the first three eigenmodes of the estimated covariance. Recalling that our toy model foregrounds can be described entirely by the first eigenmode, this 
method intentionally projects out modes that are EoR-dominated by replacing all but the three highest weights in the 
eigenspectrum with $1$'s (equal weights). Again, avoiding the over-fitting of `weak' modes which are coupled to the data results in negligible signal loss. However, we do 
not perfectly recover the shape of EoR because we lost information when projecting out certain modes. While this case is illuminating for the toy model, in practice it is not obvious which eigenmodes are FG vs. EoR dominated, so selecting a subset of modes to down-weight is not trivial.

The last regularization scheme we are highlighting here is setting $\widehat{\textbf{C}} = \widehat{\textbf{C}} \circ \textbf{I}$ (element-wise multiplication), or inverse variance weighting (keeping only the diagonal elements of $\widehat{\textbf{C}}$). In the bottom right 
panel of Figure \ref{fig:toy_sigloss8}, we see that this method does not down-weight the foregrounds --- this regularization altered $\widehat{\textbf{C}}$ in a way where it's no longer coupled to both the `strong' or `weak' eigenmodes. For this toy model, 
our foregrounds are spread out in frequency and therefore have non-negligible frequency-frequency correlations. Multiplying by 
the identity, element-wise, results in a diagonal matrix, meaning we are only left with correlation information between the same 
two frequencies. Because we disregard information from all other frequency combinations in this case, we do a poor job 
suppressing the foreground. But because we de-coupled the whole eigenspectrum from the data, we also avoid signal loss. Although this method did not successfully recover EoR for this particular simulation, it is important that we show that there 
are many options for estimating a covariance matrix, and some may down-weight certain eigenmodes more than others based on the spectral nature 
of the components in a dataset. 
%One may imagine a situation where a particular systematic is contained to an isolated 
%frequency (such as radio frequency interference or crosstalk). In such a case, preserving only the diagonal elements of $
%\widehat{\textbf{C}}$ would be an effective way of removing this contamination.
